Synthesising Concise Termination and Non-Termination Proofs for Bit-Vector Programs

### 2014-07-07 papers/termination/paper.tex
Change title
Edits, Skolemization

\title{Synthesising Termination Proofs for Bit-Vector Programs}

@@

We present a reduction from program \emph{termination} to program
\emph{synthesis} via a fragment of second-order logic.  This reduction
allows us to generate nonlinear, lexicographic ranking functions and
nonlinear recurrence sets that are correct for fixed-width machine arithmetic
and floating-point arithmetic.  We use this reduction to build a sound and
complete procedure for the termination analysis of programs with fixed-width
integers and IEEE floating-point arithmetic.
%
\end{abstract}

@@

Proving program termination is typically done by finding a \emph{ranking
function} for the program states, i.e.  a monotone map from the program's
state space to a well-ordered set.  Historically, the search for ranking
functions has been constrained in various syntactic ways, leading to
incompleteness, and is performed over abstractions that do not soundly
capture the behaviour of physical computers.  In this paper, we present a
sound and complete method for deciding whether a program with a fixed amount of
storage terminates.

When surveying the area of program termination chronologically, we observe
an initial focus on monolithic approaches based on a single measure shown to
decrease over all program
paths~\cite{DBLP:conf/vmcai/P04,DBLP:conf/cav/BradleyMS05}, followed by more
recent techniques that use termination arguments based on Ramsey's
theorem~\cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}.
The latter proof style builds an argument that a transition relation is disjunctively well founded
by composing several small well-foundedness arguments.
The main benefit of this approach is
the simplicity of local termination measures in contrast to global ones. 
For instance, there are cases in which linear arithmetic suffices when using
local measures, while corresponding global measures require non-linear
functions or lexicographic orders.

@@

As Ramsey-based approaches are limited by the state of the art in safety
checking, recent research shifts back to more complex termination arguments
that are easier to
check~\cite{DBLP:conf/cav/KroeningSTW10,DBLP:conf/tacas/CookSZ13}. 
Following the same trend, we investigate its extreme: \emph{unrestricted}
termination arguments.  This means that our ranking functions may involve
non-linearity and lexicographic orders: we do not commit to any particular
syntactic form, and do not use templates.  Furthermore, our approach allows
us to simultaneously search for proofs of non-termination, which take the form
of recurrence sets.

@@

% we propose a general framework that uniformly computes lexicographic,
% non-linear ranking functions supported by non-linear inductive invariants
% for loops with arbitrary guards and transitions over bit-vectors and floats. 
% it also computes non-linear recurrence sets, which comprise non-termination
% proofs.  our encoding treats multiple loops and nested loops uniformly,
% without having to enumerate lassos.

Our approach is to phrase the termination and non-termination problems as
second-order satisfaction.  We identify a fragment of second-order
logic that is expressive enough to capture both problems, while being
restrictive enough to admit effective solvers.  Subsequently, we propose a method for
checking the satisfiability of a formula in this fragment by
solving an isomorphic program synthesis problem
(Section~\ref{sec:synthesis}).  Our method is sound and
complete for deterministic bit-vector programs -- for any program, we will find
either a proof of its termination or non-termination.  For non-deterministic
programs, we are only semi-complete -- if the program terminates, we will find a proof,
but if it does not terminate we may fail to discover a recurrence set.

@@

\begin{figure*}
\centering
 \begin{tabular}{|ll||c|c|c|c|c|c|c|c|}
 \hline
  & & \multicolumn{8}{c|}{Program} \\
  & & \multicolumn{2}{c|}{Rationals/Integers} & \multicolumn{2}{c|}{Reals} & \multicolumn{2}{c|}{Bit-vectors} & \multicolumn{2}{c|}{Floats} \\
  & & L & NL & L & NL & L & NL & L & NL \\
  \hline
  \hline
  \multirow{4}{*}{Ranking} & Linear lexicographic &  \cite{DBLP:conf/popl/Ben-AmramG13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13,DBLP:conf/vmcai/P04} & - & \cite{DBLP:conf/tacas/LeikeH14} & - &\checkmark&\checkmark&\checkmark&\checkmark\\
   & Linear non-lexicographic & \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10} & \cite{DBLP:conf/vmcai/BradleyMS05} & \cite{DBLP:conf/tacas/LeikeH14} & - & \checkmark~ \cite{DBLP:conf/tacas/CookKRW10} &\checkmark~ \cite{DBLP:conf/tacas/CookKRW10}&\checkmark&\checkmark\\
   & Non-linear lexicographic & - & - & - & - &\checkmark&\checkmark&\checkmark&\checkmark\\
   & Non-linear non-lexicographic & \cite{DBLP:conf/vmcai/BradleyMS05} &  \cite{DBLP:conf/vmcai/BradleyMS05} & - & - &\checkmark&\checkmark&\checkmark&\checkmark\\
   \hline
 \end{tabular}

 \caption{Summary of related termination analyses. Legend: \checkmark = we can handle; - = no available works; L = linear; NL = non-linear.} \label{fig:handletable}
\end{figure*}

@@

\item We designed a bit-level accurate technique for computing ranking
functions and recurrence sets that correctly accounts for the wrap-around
behavior caused by under- and overflows in bit-vector and floating-point arithmetic.  Our
technique is not restricted to finding linear ranking functions, but can
also compute lexicographic non-linear ones.

\item  We rephrased the termination and non-termination problems as
second-order $\exists \forall$ satisfaction problems.  This formulation
captures the (non-)termination
properties of all of the loops in the program simultaneously, including
nested loops.
% In contrast to previous approaches to program synthesis, we
% found that using genetic programming was very effective at generating
% candidate termination and non-termination proofs.

@@

\item Program (b) breaks assumption (ii), in that it has no linear ranking
function.  We prove that this loop terminates by finding the non-linear
ranking function $R(x) = |x|$.

@@

\item Program (g) is a linear program that is shown
in~\cite{DBLP:conf/tacas/CookSZ13} not to admit (without prior manipulation)
a lexicographic linear ranking function.  With our technique we can find the
non-linear ranking function $R(x) = |x|$.

@@

In our setting this type of programs does not need special treatment, as we
can find a non-linear lexicographic ranking function $R(x, y, z) = (x < y,
z)$.\footnote{This termination argument is somewhat subtle.  The Boolean
values $\mathit{false}$ and $\mathit{true}$ are interpreted as 0 and 1,
respectively.  The Boolean $x < y$ thus eventually decreases, that is to say
once a state with $x \geq y$ is reached, $x$ never again becomes greater
than $y$.  This means that as soon as the ``else'' branch of the if
statement is taken, it will continue to be taken in each subsequent
iteration of the loop.  Meanwhile, if $x < y$ has not decreased (i.e., we
have stayed in the same branch of the ``if''), then $z$ does decrease. 
Since a Boolean only has two possible values, it cannot decrease
indefinitely.  Since $z > 0$ is a conjunct of the loop guard, $z$ cannot
decrease indefinitely, and so $R$ proves that the loop is well founded.}

@@

As well as machine arithmetic differing from Peano arithmetic on the
operators they have in common, computers have several ``bitwise'' operations
that are not taken as primitive in the theory of integers.  These operations
include the Boolean operators \texttt{and, or, not, xor} applied to each
element of the bit vector.  Computer programs often make use of these
operators, which are non-linear when interpreted in the standard model of
Peano arithmetic\footnote{Some of these operators can be seen as
linear in a different algebraic structure, e.g.~\texttt{xor} corresponds to
addition in the Galois field $\mathrm{GF}(2^k)$.}.

@@

Due to its expressiveness, second-order logic is very difficult to reason
about, with many second-order theories becoming undecidable even when the
corresponding first-order theory is decidable.
To make solutions of our constraints tractable, we identify a fragment of second-order logic
with a constrained use of quantification that is expressive enough to encode
both termination and non-termination, while admitting a semi-decision procedure.
We will suggestively refer to the fragment as the \emph{synthesis fragment}:

@@

We will begin our discussion by showing how to encode in the synthesis fragment the
\mbox{(non-)termination} of a program consisting of a single loop with no nesting.
For the time being, a loop $L(G, T)$ is defined by a predicate $G$ and relation $T$.
States $x$ satisfying the loop's guard are given by the
predicate $G(x)$.  The body of the loop is encoded as the transition
relation $T(x, x')$, meaning that state $x'$ is reachable from state $x$ via
a single iteration of the loop body.  For example, the loop in
Figure~\ref{fig:motivation.a} is encoded as:
%
\begin{align*}
G(x) & = \{ x \mid x>0 \} \\
T(x,x') &= \{ \langle x, x' \rangle \mid x' = (x - 1) \, \& \, x \}
\end{align*}
We will abbreviate this with the notation:
\begin{align*}
G(x) & \triangleq x > 0 \\
T(x, x') & \triangleq x' = (x - 1) \, \& \, x
\end{align*}

@@

\begin{definition}[Non-Termination Formula -- Closed Recurrence Set {\bf [CNT]}]
\label{def:cnt}
 \begin{align*}
  \exists N, x_0 . \forall x, x' . & N(x_0) ~ \wedge \\ & N(x) \rightarrow G(x) ~ \wedge \\
							& N(x) \wedge T(x, x') \rightarrow N(x') 
 \end{align*}
\end{definition}

@@

\noindent {\bf Unconditional termination.}
To show that $L(G, T)$ is unconditionally terminating (i.e. it eventually terminates regardless of the
state it starts in), it suffices to find a ranking function for \mbox{$T \cap (G \times X)$},
i.e. $T$ restricted to states satisfying the loop's guard.
% If $L$ contains nested loops, we replace $T$ with an over-approximation of the loop body $T_o$,
% as we will discuss in Section~\ref{sec:env}.

@@

Since our program's state space is finite, a transition relation
induces a infinite execution iff some state is visited infinitely
often, or equivalently $ \exists x . T^+(x, x)$.
Deciding satisfiability of this formula directly would require a logic
that includes a transitive closure operator, $T^+$.  Rather than
introduce such an operator, we will characterise non-termination
using a synthesis formula.

Gupta et al.~\cite{DBLP:conf/popl/GuptaHMRX08} characterize non-termination
of a transition relation $T$ by the existence of an \emph{(open) recurrence
set}, i.e.  a nonempty set of states $N$ such that for each $s \in N$ there
exists a transition to some $s' \in N$.
The notion of open recurrence set is encoded by formula {\bf [ONT]} in Definition~\ref{def:ont}.
% Dual to the termination problem, when formulating $L$'s non-termination, we only need to consider an under-approximation of its transition
% relation~$T_u$. 
% Again, this distinction is mostly important when dealing with nested loops,
% as illustrated in Section~\ref{sec:env}.

\begin{theorem}
\label{thm:ont}
 The loop $L(G, T)$ has an infinite execution iff Defintion~\ref{def:ont} is satisfiable.
\end{theorem}

@@

\noindent {\bf Conditional termination.}
Given a loop $L=(G,T_o)$, where $T_o$ is an over-approximation of its transition relation, 
if $L$'s termination depends on the state it begins
executing in, we say that $L$ is \emph{conditionally terminating}.
The information we require of the environment is a predicate $I$ which
over-approximates the set of states that $L$ may begin executing in.
That is to say, for each state $x$ that is reachable on entry to $L$,
we have $I(x)$.

@@

Therefore the number of non-linear functions $f$ with $K(f) \leq \kappa$
grows exponentially with $\kappa$, while the number of linear functions
grows only linearly.  This shows that assuming the existence of a linear
ranking function is much stronger than assuming the existence of a short
termination proof.  In addition to being a much weaker assumption,
we argue that low Kolmogorov complexity is a more natural assumption than
linearity.  Humans tend to write programs they can understand.  It's hard
to quantify exactly what ``understandable'' means, but we think that
``having a short proof'' is closer to the mark than ``having a linear proof''.

@@

While the synthesis of termination arguments for linear programs over
integers is indeed well covered in the literature, there is very limited
work for programs over machine integers.  Cook et al.~present a method
based on a reduction to Presburger arithmetic, and a template-matching
approach for predefined classes of ranking functions based on reduction to
SAT- and QBF-solving~\cite{DBLP:conf/tacas/CookKRW10}.  Similarly, the only
work we are aware of that can compute non-linear ranking functions for
imperative loops with polynomial guards and polynomial assignments
is~\cite{DBLP:conf/vmcai/BradleyMS05}.  However, this work extends only to
polynomials.

>>>>

\title{Synthesising Concise Termination and Non-Termination Proofs for Bit-Vector Programs}

@@

We present a reduction from program \emph{termination} to program
\emph{synthesis} via a fragment of second-order logic.  This reduction
allows us to generate nonlinear, lexicographic ranking functions and
nonlinear recurrence sets that are correct for fixed-width machine arithmetic
and floating-point arithmetic.  We use this reduction to build a sound and
complete procedure for the termination analysis of finite state programs with fixed-width
integers and IEEE floating-point arithmetic.
%
\end{abstract}

@@

Proving program termination is typically done by finding a \emph{ranking
function} for the program states, i.e.  a monotone map from the program's
state space to a well-ordered set.  Historically, the search for ranking
functions has been constrained in various syntactic ways, leading to
incompleteness, and is performed over abstractions that do not soundly
capture the behaviour of physical computers.  In this paper, we present a
sound and complete method for deciding whether a program with a fixed amount of
storage terminates.  Since such programs are necessarily finite state, our
problem is much easier than Turing's, but is more relevant to analysing computer
programs.

When surveying the area of program termination chronologically, we observe
an initial focus on monolithic approaches based on a single measure shown to
decrease over all program
paths~\cite{DBLP:conf/vmcai/P04,DBLP:conf/cav/BradleyMS05}, followed by more
recent techniques that use termination arguments based on Ramsey's
theorem~\cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}.
The latter proof style builds an argument that a transition relation is disjunctively well founded
by composing several small well-foundedness arguments.
The main benefit of this approach is
the simplicity of local termination measures in contrast to global ones. 
For instance, there are cases in which linear arithmetic suffices when using
local measures, while corresponding global measures require nonlinear
functions or lexicographic orders.

@@

As Ramsey-based approaches are limited by the state of the art in safety
checking, recent research shifts back to more complex termination arguments
that are easier to
check~\cite{DBLP:conf/cav/KroeningSTW10,DBLP:conf/tacas/CookSZ13}. 
Following the same trend, we investigate its extreme: \emph{unrestricted}
termination arguments.  This means that our ranking functions may involve
nonlinearity and lexicographic orders: we do not commit to any particular
syntactic form, and do not use templates.  Furthermore, our approach allows
us to simultaneously search for proofs of non-termination, which take the form
of recurrence sets.

@@

% we propose a general framework that uniformly computes lexicographic,
% nonlinear ranking functions supported by nonlinear inductive invariants
% for loops with arbitrary guards and transitions over bit-vectors and floats. 
% it also computes nonlinear recurrence sets, which comprise non-termination
% proofs.  our encoding treats multiple loops and nested loops uniformly,
% without having to enumerate lassos.

Our approach is to phrase the termination and non-termination problems as
second-order satisfaction.  We identify a fragment of second-order
logic that is expressive enough to capture both problems, while being
restrictive enough to admit effective solvers.  Subsequently, we propose a method for
checking the satisfiability of a formula in this fragment by
solving an isomorphic program synthesis problem
(Section~\ref{sec:synthesis}).  Our method is sound and
complete for bit-vector programs -- for any program, we will find
a proof of either its termination or non-termination.

@@

\begin{figure*}
\centering
 \begin{tabular}{|ll||c|c|c|c|c|c|c|c|}
 \hline
  & & \multicolumn{8}{c|}{Program} \\
  & & \multicolumn{2}{c|}{Rationals/Integers} & \multicolumn{2}{c|}{Reals} & \multicolumn{2}{c|}{Bit-vectors} & \multicolumn{2}{c|}{Floats} \\
  & & L & NL & L & NL & L & NL & L & NL \\
  \hline
  \hline
  \multirow{4}{*}{Ranking} & Linear lexicographic &  \cite{DBLP:conf/popl/Ben-AmramG13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13,DBLP:conf/vmcai/P04} & - & \cite{DBLP:conf/tacas/LeikeH14} & - &\checkmark&\checkmark&\checkmark&\checkmark\\
   & Linear non-lexicographic & \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10} & \cite{DBLP:conf/vmcai/BradleyMS05} & \cite{DBLP:conf/tacas/LeikeH14} & - & \checkmark~ \cite{DBLP:conf/tacas/CookKRW10} &\checkmark~ \cite{DBLP:conf/tacas/CookKRW10}&\checkmark&\checkmark\\
   & Nonlinear lexicographic & - & - & - & - &\checkmark&\checkmark&\checkmark&\checkmark\\
   & Nonlinear non-lexicographic & \cite{DBLP:conf/vmcai/BradleyMS05} &  \cite{DBLP:conf/vmcai/BradleyMS05} & - & - &\checkmark&\checkmark&\checkmark&\checkmark\\
   \hline
 \end{tabular}

 \caption{Summary of related termination analyses. Legend: \checkmark = we can handle; - = no available works; L = linear; NL = nonlinear.} \label{fig:handletable}
\end{figure*}

@@

\item We designed a bit-level accurate technique for computing ranking
functions and recurrence sets that correctly accounts for the wrap-around
behavior caused by under- and overflows in bit-vector and floating-point arithmetic.  Our
technique is not restricted to finding linear ranking functions, but can
also compute lexicographic nonlinear ones.

\item  We rephrased the termination and non-termination problems as
second-order $\exists \forall$ satisfaction problems.  This formulation
captures the (non-)termination
properties of all of the loops in the program, including
nested loops.  We can use this to analyse all the loops at once,
or one at a time.
% In contrast to previous approaches to program synthesis, we
% found that using genetic programming was very effective at generating
% candidate termination and non-termination proofs.

@@

\item Program (b) breaks assumption (ii), in that it has no linear ranking
function.  We prove that this loop terminates by finding the nonlinear
ranking function $R(x) = |x|$.

@@

\item Program (g) is a linear program that is shown
in~\cite{DBLP:conf/tacas/CookSZ13} not to admit (without prior manipulation)
a lexicographic linear ranking function.  With our technique we can find the
nonlinear ranking function $R(x) = |x|$.

@@

In our setting this type of programs does not need special treatment, as we
can find a nonlinear lexicographic ranking function $R(x, y, z) = (x < y,
z)$.\footnote{This termination argument is somewhat subtle.  The Boolean
values $\mathit{false}$ and $\mathit{true}$ are interpreted as 0 and 1,
respectively.  The Boolean $x < y$ thus eventually decreases, that is to say
once a state with $x \geq y$ is reached, $x$ never again becomes greater
than $y$.  This means that as soon as the ``else'' branch of the if
statement is taken, it will continue to be taken in each subsequent
iteration of the loop.  Meanwhile, if $x < y$ has not decreased (i.e., we
have stayed in the same branch of the ``if''), then $z$ does decrease. 
Since a Boolean only has two possible values, it cannot decrease
indefinitely.  Since $z > 0$ is a conjunct of the loop guard, $z$ cannot
decrease indefinitely, and so $R$ proves that the loop is well founded.}

@@

As well as machine arithmetic differing from Peano arithmetic on the
operators they have in common, computers have several ``bitwise'' operations
that are not taken as primitive in the theory of integers.  These operations
include the Boolean operators \texttt{and, or, not, xor} applied to each
element of the bit vector.  Computer programs often make use of these
operators, which are nonlinear when interpreted in the standard model of
Peano arithmetic\footnote{Some of these operators can be seen as
linear in a different algebraic structure, e.g.~\texttt{xor} corresponds to
addition in the Galois field $\mathrm{GF}(2^k)$.}.

@@

Due to its expressiveness, second-order logic is very difficult to reason
about, with many second-order theories becoming undecidable even when the
corresponding first-order theory is decidable.
To make solving our constraints tractable, we identify a fragment of second-order logic
with a constrained use of quantification that is expressive enough to encode
both termination and non-termination of finite programs, while admitting a semi-decision procedure.
We will suggestively refer to the fragment as the \emph{synthesis fragment}:

@@

We will begin our discussion by showing how to encode in the synthesis fragment the
\mbox{(non-)termination} of a program consisting of a single loop with no nesting.
For the time being, a loop $L(G, T)$ is defined by its guard $G$ and body $T$
such that states $x$ satisfying the loop's guard are given by the
predicate $G(x)$.  The body of the loop is encoded as the transition
relation $T(x, x')$, meaning that state $x'$ is reachable from state $x$ via
a single iteration of the loop body.  For example, the loop in
Figure~\ref{fig:motivation.a} is encoded as:
%
\begin{align*}
G(x) & = \{ x \mid x>0 \} \\
T(x,x') &= \{ \langle x, x' \rangle \mid x' = (x - 1) \, \& \, x \}
\end{align*}
We will abbreviate this with the notation:
\begin{align*}
G(x) & \triangleq x > 0 \\
T(x, x') & \triangleq x' = (x - 1) \, \& \, x
\end{align*}

@@

\begin{definition}[Non-Termination Formula -- Closed Recurrence Set {\bf [CNT]}]
\label{def:cnt}
 \begin{align*}
  \exists N, x_0 . \forall x, x' . & N(x_0) ~ \wedge \\ & N(x) \rightarrow G(x) ~ \wedge \\
							& N(x) \wedge T(x, x') \rightarrow N(x') 
 \end{align*}

\begin{definition}[Non-Termination Formula -- Skolemized Open Recurrence Set  {\bf [SNT]}]
\label{def:ont}
 \begin{align*}
  \exists N, S, x_0 . \forall x . & N(x_0) ~\wedge \\ &  N(x) \rightarrow G(x) ~ \wedge \\
							& N(x) \rightarrow T(x, S(x)) \wedge N(S(x))
 \end{align*}
\end{definition}

@@

\noindent {\bf Unconditional termination.}
We say that a loop $L(G, T)$ is unconditionally terminating iff it eventually
terminates regardless of the state it starts in. To prove unconditional termination, it suffices
to find a ranking function for \mbox{$T \cap (G \times X)$}, i.e. $T$ restricted to states satisfying the loop's guard.
% If $L$ contains nested loops, we replace $T$ with an over-approximation of the loop body $T_o$,
% as we will discuss in Section~\ref{sec:env}.

@@

Since our program's state space is finite, a transition relation
induces an infinite execution iff some state is visited infinitely
often, or equivalently $ \exists x . T^+(x, x)$.
Deciding satisfiability of this formula directly would require a logic
that includes a transitive closure operator, $\bullet^+$.  Rather than
introduce such an operator, we will characterise non-termination
using a synthesis formula.

Gupta et al.~\cite{DBLP:conf/popl/GuptaHMRX08} characterise non-termination
of a transition relation $T$ by the existence of an \emph{(open) recurrence
set}, i.e.  a nonempty set of states $N$ such that for each $s \in N$ there
exists a transition to some $s' \in N$.
The notion of open recurrence set is encoded by formula {\bf [ONT]} in Definition~\ref{def:ont}.
% Dual to the termination problem, when formulating $L$'s non-termination, we only need to consider an under-approximation of its transition
% relation~$T_u$. 
% Again, this distinction is mostly important when dealing with nested loops,
% as illustrated in Section~\ref{sec:env}.

\begin{theorem}
\label{thm:ont}
 The loop $L(G, T)$ has an infinite execution iff formula {\bf [ONT]} (Definition~\ref{def:ont}) is satisfiable.
\end{theorem}

@@

\noindent {\bf Conditional termination.}
Given a loop $L(G,T_o)$, where $T_o$ is an over-approximation of its transition relation, 
if $L$'s termination depends on the state it begins
executing in, we say that $L$ is \emph{conditionally terminating}.
The information we require of the environment is a predicate $I$ which
over-approximates the set of states that $L$ may begin executing in.
That is to say, for each state $x$ that is reachable on entry to $L$,
we have $I(x)$.

@@

Therefore the number of nonlinear functions $f$ with $K(f) \leq \kappa$
grows exponentially with $\kappa$, while the number of linear functions
grows only linearly.  This shows that assuming the existence of a linear
ranking function is much stronger than assuming the existence of a short
termination proof.  In addition to being a much weaker assumption,
we argue that low Kolmogorov complexity is a more natural assumption than
linearity.  Humans tend to write programs they can understand.  It's hard
to quantify exactly what ``understandable'' means, but we think that
``having a short proof'' is closer to the mark than ``having a linear proof''.

@@

While the synthesis of termination arguments for linear programs over
integers is indeed well covered in the literature, there is very limited
work for programs over machine integers.  Cook et al.~present a method
based on a reduction to Presburger arithmetic, and a template-matching
approach for predefined classes of ranking functions based on reduction to
SAT- and QBF-solving~\cite{DBLP:conf/tacas/CookKRW10}.  Similarly, the only
work we are aware of that can compute nonlinear ranking functions for
imperative loops with polynomial guards and polynomial assignments
is~\cite{DBLP:conf/vmcai/BradleyMS05}.  However, this work extends only to
polynomials.
